---
title: "CS&SS 569 Visualizing Data and Models"
subtitle: "Lab 5: Intro to `tile`"
author: "Brian Leung"
institute: "Department of Political Science, UW"
date: \today
output:
  beamer_presentation:
    incremental: yes
link-citations: yes
linkcolor: blue
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 
- Overview of `tile`
- Preview of three examples
  - Scatterplot: HW1 example
  - Expected probabilities and first differences: Voting example
  - Ropeladder: Crime example
- Installing `tile` and `simcf`
- Walking through examples

## Overview of `tile`
- A fully featured R graphics package built on the `grid` graphics environment
- Features:
  - Make standard displays like scatterplots, lineplots, and dotplots
  - Create more experimental formats like ropeladders
  - Summarize uncertainty in inferences from model
  - Avoid extrapolation from the original data underlying your model
  - Fully control titles, annotation, and layering of graphical elements
  - Build your own tiled graphics from primitives
- Work well in combination with `simcf` package
   - Calculate counterfactual expected values, first differences, and relative risks, and their confidence intervals
   - More later

## Overview of `tile`
- Three steps to make `tile` plots (from Chris's [“Tufte Without Tears”](https://faculty.washington.edu/cadolph/software/tilepres4.pdf))
  1. **Create data traces**: Each trace contains the data and graphical parameters needed to plot a single set of graphical elements to one or more plots
      - Could be a set of points, or text labels, or lines, or a polygon
      - Could be a set of points and symbols, colors, labels, fit line, CIs, and/or
  extrapolation limits
      - Could be the data for a dotchart, with labels for each line
      - Could be the marginal data for a rug
      - All annotation must happen in this step
      - Basic traces: `linesTile()`, `pointsile()`, `polygonTile()`, `polylinesTile()`, and `textTile()`
      - Complex traces: `lineplot()`, `scatter()`, `ropeladder()`, and `rugTile()`
  
## Overview of `tile`
>  - Primitive trace functions: 
>    - `linesTile()`: Plot a set of connected line segments
>    - `pointsTile()`: Plot a set of points
>    - `polygonTile()`: Plot a shaded region
>    - `polylinesTile()`: Plot a set of unconnected line segments
>    - `textTile()`: Plot text labels 
>  - Complex traces for model or data exploration: 
>    - `lineplot()`: Plot lines with confidence intervals, extrapolation warnings
>    - `ropeladder()`: Plot dotplots with confidence intervals, extrapolation warnings, and shaded ranges
>    - `rugTile()`: Plot marginal data rugs to axes of plots
>    - `scatter()`: Plot scatterplots with text and symbol markers, fit lines, and confidence intervals 

## Overview of `tile`
- Three steps to make `tile` plots (from Chris's [“Tufte Without Tears”](https://faculty.washington.edu/cadolph/software/tilepres4.pdf))
  1. **Create data trace**: Each trace contains the data and graphical parameters needed to plot a single set of graphical elements to one or more plots
  2. **Plot the data traces**: Using the `tile()` function, simultaneously plot all traces to all plots
      - This is the step where the scaffolding gets made: axes and titles 
      - Set up the rows and columns of plots
      - Titles of plots, axes, rows of plots, columns of plots, etc. 
      - Set up axis limits, ticks, tick labels, logging of axes
  3. **Examine output and revise**: Look at the graph made in step 2, and tweak the input parameters for steps 1 and 2 to make a better graph

## Three examples
>  - Scatterplot: HW1 example
>  - Expected probabilities and first differences: Voting example
>  - Ropeladder: Crime examples (if time permits)
  
## Scatterplot: HW 1 example
```{r, echo=F}
knitr::include_graphics("plot/iverScatter.pdf")
```

## Expected probabilities and first differences: Voting example
```{r, echo=F}
knitr::include_graphics("plot/educationEV.pdf")
```

## Scatterplot: HW 1 example
- A quick detour to model results presentation and the logic of simulation (consult POLS/CSSS 510:MLE::[Topic 3](https://faculty.washington.edu/cadolph/mle/topic3.p.pdf))
  1. Obtain estimated parameters ($\hat{\beta}_{k}$) and standard errors (more precisely, the variance-covariance matrix)
      - `lm()`, `glm()`...; `coef()`, `vcov()`...
      - What you see in usual regression tables
  2. Capture our uncertainty around $\hat{\beta}_{k}$ by drawing, say, 10,000 $\tilde{\beta}_{k}$ from a multivariate normal distribution
      - `MASS::mvrnorm()`
  3. Specify counterfactual scenarios (hypothetical values for all relevant covariates $x_{k}$)
      - `simcf::cfMake`, `cfChange`...
  4. Simulate quantities of interest by compounding those 10,000 $\tilde{\beta}_{k}$ with counterfactual scenarios 
      - Then compute average (point estimate) and appropriate percentiles (confidence intervals)
      - `simcf::logitsimev()` for expected values for logit models
      - `logitsimfd` for first differences
      - `logitsimrr` for relative risks

## Expected probabilities and first differences: Voting example
```{r, echo=F}
knitr::include_graphics("plot/educationEV.pdf")
```

## Expected probabilities and first differences: Voting example
```{r, echo=F}
knitr::include_graphics("plot/marriedEV.pdf")
```

## Expected probabilities and first differences: Voting example
```{r, echo=F}
knitr::include_graphics("plot/marriedFDRR.pdf")
```

## Ropeladder: Crime example (if time permits)
```{r, echo=F}
knitr::include_graphics("plot/ropeladderEx0.pdf")
```

## Ropeladder: Crime example (if time permits)
```{r, echo=F}
knitr::include_graphics("plot/ropeladderEx1.pdf")
```

## 
```{r, echo=F, fig.height=0.5}
knitr::include_graphics("plot/ropeladderEx2.pdf")
```

## Ropeladder: Crime example (if time permits)
```{r, echo=F}
knitr::include_graphics("plot/ropeladderEx3.pdf")
```

##  Installing `tile` and `simcf`
- Go to Chris's website, [Software section](https://faculty.washington.edu/cadolph/index.php?page=60)
- Also download all `R` scripts and data under today's Lab section
