---
title: "CS&SS 569 Visualizing Data and Models"
subtitle: "Lab 2: Intro to \\LaTeX{} with \\texttt{R Markdown} and \\texttt{Overleaf}"
author: "Brian Leung"
institute: "Department of Political Science, UW"
date: "`r Sys.Date()`"
output: 
  beamer_presentation:
    incremental: true
link-citations: yes
linkcolor: blue
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Agenda

1. Logistics

2. `R Markdown` and HW1

3. \LaTeX{} and `Overleaf`

# Homework Submission

- **Use Canvas** not email


# `R Markdown` and HW1

- Problem 1: Attach files (PDF/picture), and offer your critique and remedy

- Problem 2: Read data and reproduce the plot
  - *Do not spend more than two hours* 

# `R Markdown`

- `R Markdown` file (`.Rmd`) offers an integrated framework
  - To contain both narrative text, code chunks and outputs such as graphs
  - To render ("knit") high quality, reproducible outputs
      - `HTML`, `PDF`, `Word`, `Beamer`, etc.
      - I write my slides using `R Markdown` (in `Beamer`: a \LaTeX{} class of presentation slides)
      - Great way to submit your homework
  - \LaTeX{} code is supported

# `R Markdown` and `TinyTeX`

- To compile a `R Markdown` document to `PDF`, you need to install \LaTeX{}
  - If you haven't installed any previous \LaTeX{} distribution, I recommend `TinyTeX`

# `R Markdown` and `TinyTeX`

- "`TinyTeX` is a lightweight, portable, cross-platform, and easy-to-maintain LaTeX distribution":  

```{r, eval = FALSE}
install.packages('tinytex')
tinytex::install_tinytex()
```

# `R Markdown` in practices: two examples 

- `lab2_RMarkdownSample.Rmd`: cover most of the basic functionalities in R Markdown; good for future reference

- `CSSS569HW1Start.Rmd`: a template to get started with HW 1

# Break 

# Intro to \TeX{} and \LaTeX{}

- \TeX{} is a *typesetting engine* designed by Donald Knuth, a computer scientist and mathematician at Stanford
  - For typesetting scientific text and mathematical formulas
  - Modern extensions of the \TeX{} engines include `pdfTeX`, `XeTeX`, `LuaTeX`, etc.

# Intro to \TeX{} and \LaTeX{}

- \LaTeX{} is a *document preparation system, or a macro package*, built on top of the \TeX{} engine, with features:
  - Typesetting journal articles, technical reports, books, and slides
  - Control over large documents containing sectioning, cross-references, tables and figures
  - Typesetting of complex mathematical formulas
  - Advanced typesetting of mathematics with AMS-LaTeX
  - Automatic generation of bibliographies and indexes
  - Multi-lingual typesetting
  - See more \href{https://www.latex-project.org}{here}

# Intro to \TeX{} and \LaTeX{}
- Popular *implementations, or distributions*, of \TeX{}/\LaTeX{}
  - MacTeX for Mac OS: <http://www.tug.org/mactex/>
  - MiKTeX for Windows: <https://miktex.org>

# Intro to \TeX{} and \LaTeX{}
- \LaTeX{} vs. other word processors (e.g. Microsoft Word)
  - Microsoft Word/Power Point
    - WYSIWYG: What You See Is What You Get
    - You interact with a user interface to control the document layout while typing text 
    - What is displayed on the screen resembles what will be printed 
  - \LaTeX{}
    - You provide "\LaTeX{} commands" to specify the layout, structure, and details of the document:
    - `\command[optional parameter]{parameter}`
    - And *typeset* the document using the \TeX{} engine and compile the output
    
# Intro to \TeX{} and \LaTeX{}
- The input for \LaTeX{} is a plain text file (.tex)
  - You need a text editor!
- Numerous popular text editors
  - Specific: Texmaker, TeXShop, TeXstudio, TeXworks...
  - Generic: Emacs (Aquamacs), Vim, Sublime, Atom...


# Intro to \LaTeX{} with `Overleaf`
- All the above sound pretty complicated...
- `Overleaf`: <https://www.overleaf.com/>
  - An online \LaTeX{} editor
    - Integrated PDF preview pane
    - Quality of life features: auto-complete commands, auto-close brackets, keyboard shortcuts, etc.
    - Numerous templates: journal articles, books, CVs, slides, posters, etc.
    - Easy collaboration (But not free) 
    - Integrated with `Zotero` and `Mendeley` for bibliography management
    - Integrated with `Git` for version control
    
# Intro to \LaTeX{} with `Overleaf`
- Before we dive in, useful resources
  - \href{https://tobi.oetiker.ch/lshort/lshort.pdf}{The Not So Short Introduction to \LaTeXe{} (Oetiker et al., 2018)}
    - Learn \LaTeX{} in 139 pages / minutes
  - \href{https://www.overleaf.com/learn/}{`Overleaf` documentation}
    - Contains intro to basic \LaTeX{}, `Overleaf`, and many practical guides
  - \href{https://tex.stackexchange.com}{\TeX{} at StackExchange}
  -  General: \href{https://en.wikibooks.org/wiki/LaTeX/Mathematics}{Mathematics} and \href{https://en.wikibooks.org/wiki/LaTeX/Tables}{Tables} and \href{https://sites.google.com/site/kochiuyu/Tikz}{TikZ}
  -  Beamer Theme: \href{https://hartwork.org/beamer-theme-matrix/}{here}
  -  Bibliography: \href{http://merkel.texture.rocks/Latex/natbib.php}{natbib}, \href{https://www.doi2bib.org/}{doi2bib}, \href{https://text2bib.economics.utoronto.ca/}{text2bib}
  - Other: \href{https://gking.harvard.edu/pages/student-materials}{here}

  
# Intro to \LaTeX{} with `Overleaf`
- Some useful templates:
  - Thesis: \href{https://www.overleaf.com/latex/templates/phd-thesis-and-dissertation-latex-templates-for-harvard-princeton-and-new-york-university-nyu/wkzvtymyxqvv}{here}
  - Working paper: \href{https://www.overleaf.com/project/5e42e4a9ae15f200014a4919}{Chris's sample}
  - Academic journal: \href{https://www.overleaf.com/latex/templates/american-political-science-review-apsr-submission-template/fxffppspqczt}{here}
  - Presentation slides (Beamer): \href{https://www.overleaf.com/latex/templates/metropolis-beamer-theme/qzyvdhrntfmr}{here} and 
  \href{https://www.overleaf.com/latex/templates/presentation-template/ycwnkvzxyzwv}{here} 
  - Poster: \href{https://www.overleaf.com/latex/templates/landscape-beamer-poster-template/vjpmsxxdvtqk}{here}
  - CV: \href{https://ja.overleaf.com/latex/templates/academic-cv-template/xykxmjmkxxjm}{here} and 
  \href{https://ja.overleaf.com/gallery/tagged/cv}{here}
  - Graphs, trees, diagrams (TikZ): \href{https://www.overleaf.com/latex/examples/branching-arrows-with-decision-option-in-flowchart/txjdyggmtbjw}{here} and \href{https://www.overleaf.com/latex/examples/tikz-flowchart-of-an-epidemiology-study/hdpqksvwkqdm}{here} 











