---
title: "CS&SS 569 Visualizing Data and Models"
subtitle: "Lab 3: ggplot2"
author: "Brian Leung"
institute: "Department of Political Science, UW"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 12, fig.height = 9, message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(xaringan)
library(tidyverse)
library(scales)
library(ggrepel)
library(RColorBrewer)
```

## Introduction 

- Unpack the design philosophy and inner working of ggplot2

--

- Gain finer control over *each and every* aspect of the graphic design, rather than relying on default 

--

- Showcase some tricks and useful functions common in my own viz workflow

--

- Hopefully it benefits beginners, and intermediate (at least marginally), users to ggplot2

---

## ggplot2 design philosophy 

- Grammar of graphics 

--

  - A statistical graphic is a `mapping` of `data` variables to `aes`thetic attributes of `geom`etric objects. (Wilkinson 2005)

--

- *What* data do you want to visualize?
  
--

  - `ggplot(data = ...)`
  
--
  
- *How* are variables mapped to specific aesthetic attributes?

--

  - `aes(... = ...)`
  
--

  - positions (`x`, `y`), `shape`, `colour`, `size`, `fill`, `alpha`, `linetype`, `label`... 
  
--

  - If the value of an attribute do not vary w.r.t. some variable, don't wrap it within `aes(...)`
  
--

- *Which* geometric shapes do you use to represent the data?

--

  - `geom_{...}`: `geom_point`, `geom_line`, `geom_ribbon`, `geom_polygon`...
  
---

## ggplot2: A layered grammar

- ggplot2: A *layered* grammar of graphics (Wickham 2009)

--
  
  - Build a graphic from multiple layers; each consists of some geometric objects or transformation 
  
--

  - Use `+` to stack up layers

--

- When you declare `ggplot(data = ..., aes(...))`, you're declaring the default specification 

--

- Within each `geom_{}` layer that follows, two things are *inherited* from the default specification at the master level

--

  - Data: inherited from the default data you supply 

--
  
  - Aesthetics: inherited from the default aesthetics (`inherit.aes = TRUE`) 

--
  
  - They are convenient but can create unintended consequences

--

  - We will learn how to overwrite this property for extra flexibility 

---

## Quick demo: inherit.aes = TRUE property

```{r}
library(tidyverse)

iver_data <- read_csv("https://faculty.washington.edu/cadolph/vis/iverRevised.csv")

head(iver_data)
```

---

`r chunk_reveal("iver1", break_type = "non_seq", title = "## Why are there multiple smooth curves? -- inherit.aes")`

```{r "iver1", include = FALSE}
ggplot(iver_data, 
    aes(x = effectiveParties, 
        y = povertyReduction, 
        color = partySystem)) + #BREAK
  geom_point(
    inherit.aes = TRUE,       #BREAK2
    aes(x = effectiveParties, #BREAK2
        y = povertyReduction, #BREAK2
        color = partySystem)  #BREAK2
  ) +
  geom_smooth(
    inherit.aes = TRUE,       #BREAK3
    aes(x = effectiveParties, #BREAK3
        y = povertyReduction, #BREAK3
        color = partySystem), #BREAK3
    method = MASS::rlm        
  )
```

---

`r chunk_reveal("iver2", break_type = "non_seq", title = "## You can either overwrite the geom_smooth() layer")`


```{r "iver2", include = FALSE}
ggplot(iver_data, 
    aes(x = effectiveParties, 
        y = povertyReduction, 
        color = partySystem)) + 
  geom_point(
  ) + 
  geom_smooth(
    aes(group = 1),    #BREAK2
    method = MASS::rlm        
  )
```

---

`r chunk_reveal("iver3", break_type = "non_seq", title = "## Or localize color aesthetics to the geom_point layer()")`

```{r "iver3", include=FALSE}
ggplot(iver_data, 
    aes(x = effectiveParties, 
        y = povertyReduction)) + 
  geom_point(
    aes(color = partySystem) #BREAK2
  ) + 
  geom_smooth(
    method = MASS::rlm        
  )
```

---

## A segue to a customized ggplot2 theme 

- Chris and I wrote a customized ggplot2 theme to implement a minimalist aesthetics

--

- Can be downloaded here: https://students.washington.edu/kpleung/vis/theme_cavis.R

--

- To use it 

```{r}
source("https://students.washington.edu/kpleung/vis/theme_cavis.R")
# source(".../theme_cavis.R") if you have download and put it in your directory
```

--

- I'll use this theme throughout my vis; you don't have to (and suggestions are welcomed!)

---

`r chunk_reveal("theme_cavis", break_type = "rotate", title = "## Showcase customized theme")`

```{r "theme_cavis", include = FALSE}
ggplot(iris, 
       aes(x = Sepal.Length,
           y = Petal.Length,
           color = Species)) +
  geom_point() + 
  theme_grey() + #ROTATE 
  theme_cavis_hgrid + #ROTATE 
  theme_cavis_vgrid + #ROTATE
  theme_cavis #ROTATE
```

---

## Visualization exercises: Electric vehicles in WA state 

- Data are taken from [Washington State Department of Licensing](https://data.wa.gov/Transportation/Electric-Vehicle-Population-Data/f6w7-q2d2)

--

- "This dataset shows the Battery Electric Vehicles (BEVs) and Plug-in Hybrid Electric Vehicles (PHEVs) that are currently registered through Washington State Department of Licensing (DOL)."

--

- Load the electric vehicles data 

```{r}
ev_data <- read_csv("data/ev_data.csv")
head(ev_data)
```

---

## Visualization exercises: Electric vehicles in WA state 

- Also some auxiliary county data with number of households and their median income 

```{r}
county_data <- read_csv("data/county_data.csv")
head(county_data)
```

---

`r chunk_reveal("ev_by_county1", break_type = "auto", widths = c(90, 10), title = "## EV in WA state: Which county has most EVs?", float = "top")`

```{r "ev_by_county1", include = FALSE}
# Count EVs by county 
ev_data %>%
  count(county) %>%
  mutate(county = fct_reorder(county, n)) %>%
  arrange(desc(county)) -> 
  ev_by_county

# Load scales package for labels and breaks
library(scales)
```

---

`r chunk_reveal("ev_by_county2", break_type = "non_seq", widths = c(50, 50), title = "## EV in WA state: Which county has most EVs?")`

```{r "ev_by_county2", include = FALSE}
ggplot(ev_by_county, 
       aes(y = county, x = n)) +
  theme_cavis_vgrid +
  geom_point() +  
  geom_col(       #BREAK2
    width = 0.05, #BREAK3
    alpha = 0.5   #BREAK3
  ) +             #BREAK2
  scale_x_continuous(               #BREAK4
    trans = "log10",                #BREAK4
    breaks = 10^(0:5),              #BREAK5
    labels = scales::label_comma(), #BREAK5
  ) +                               #BREAK4
  geom_text(                                   #BREAK6
    data = slice_max(ev_by_county, n, n = 10), #BREAK8
    aes(label = n),                            #BREAK6
    nudge_x = 0.15                             #BREAK7
  ) +                                          #BREAK6
  labs(y = NULL, x = "Total number of EVs")    #BREAK9
```

---

`r chunk_reveal("ev_by_countyYear1", break_type = "auto", widths = c(90, 10), title = "## EV in WA state: Trend of number of EVs across county", float = "top")`

```{r "ev_by_countyYear1", include = FALSE}
# Count number of EVs by county
ev_data %>% 
  group_by(model_year, county) %>%
  count() ->
  ev_by_countyYear
```

---

`r chunk_reveal("ev_by_countyYear2", break_type = "auto", widths = c(50, 50), title = "## EV in WA state: Trend of number of EVs across county")`

```{r "ev_by_countyYear2", include = FALSE, eval=F}
ggplot(ev_by_countyYear,
       aes(x = model_year, 
           y = n,
           group = county,
           color = county == "King")) +
  theme_cavis_hgrid +
  geom_line() +
  scale_y_continuous(trans = "log10") +
  scale_color_manual(values = c("grey90", "red")) +
  guides(color = "none") + 
  annotate("text", x = 2018, y = 13000, 
           label = "King County", 
           color = "red") +
  geom_vline(xintercept = 2012, linetype = 2, 
             color = "grey45") +
  annotate("text", x = 2010, y = 15000, 
           label = "Tesla Model S\nintroduced", 
           color = "black") +
  labs(y = "Number of registed EVs", x = NULL)
```

---

`r chunk_reveal("price_range1", break_type = "auto", widths = c(90, 10), title = "## EV in WA state: Relationship between price and range", float = "top")`

```{r "price_range1", include = FALSE}
ev_data %>%
  distinct(make, model, model_year, ev_type, electric_range, retail_price) %>%
  filter(retail_price > 0 & electric_range > 0) ->
  ev_models

# load packages 
library(ggrepel)
library(RColorBrewer)
```

---

## A segue to choosing colors

- Use `RColorBrewer` package to choose colors

--

- Choose between sequential, diverging or qualitative palettes based on your data

--

- Check out this [website](https://colorbrewer2.org/)

---

`r chunk_reveal("color1", break_type = "auto", widths = c(50, 50), title = "## A segue to choosing colors")`

```{r "color1", include = FALSE}
RColorBrewer::display.brewer.all()
colors <- brewer.pal(n = 5, name = "Set1")
print(colors)
blue <- colors[2]
orange <- colors[5]
```

---

`r chunk_reveal("price_range2", break_type = "non_seq", widths = c(50, 50), title = "## EV in WA state: Relationship between price and range")`

```{r "price_range2", include = FALSE}
ggplot(ev_models,
       aes(x = electric_range, y = retail_price,
           color = ev_type, fill = ev_type)) +
  theme_cavis_hgrid +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm",               #BREAK2                            
              alpha = 0.15, size = 0.45) + #BREAK2                           
  scale_x_continuous(trans = "log2") + #BREAK3
  scale_y_continuous(trans = "log10",  #BREAK4
                     labels = label_dollar())+ #BREAK4
  scale_color_manual(values = c(blue, orange)) + #BREAK5
  scale_fill_manual(values = c(blue, orange)) +  #BREAK5
  geom_text_repel(                                          #BREAK6
    aes(label = paste(make, model, model_year)),            #BREAK6
    show.legend = FALSE,                                    #BREAK7
    size = 3, alpha = 0.75                                  #BREAK7
  ) +                                                       #BREAK6
  theme(legend.position = "top") +                          #BREAK8
  labs(y = "Retail price", x = "Electric range (mile)")     #BREAK9
```

---

## Another segue: how to properly save a graph? 

- Don't use the `Export` interface in your RStudio

--

- Use code for reproducibility; explicitly declare parameters (e.g. for width, height and aspect ratio) 

--

- Use `ggsave()` function for ggplot2 objects

```{r, eval = FALSE}
width <- 12
ggsave("ev_bivariate.pdf", width = width, height = width/1.618)
```

---

`r chunk_reveal("make_share_fail", break_type = "auto", widths = c(50, 50), title = "## EV in WA state: Shares of makes across counties")`

```{r "make_share_fail", include=FALSE}
# bad example
ev_data %>%
  group_by(county, make) %>%
  count() %>%
  ggplot(aes(x = n, y = make)) +
  theme_cavis_vgrid +
  geom_point() +
  facet_wrap(~ county) +
  scale_x_continuous(
    trans = "log10", 
    labels = label_log()
  )
```

---

`r chunk_reveal("make_share1", break_type = "auto", widths = c(90, 10), float = "top", title = "## EV in WA state: Shares of makes across counties")`

```{r "make_share1", include=FALSE}
# Focus on the top 7 most popular makes
ev_data %>%
  count(make) %>%
  slice_max(order_by = n, n = 7) ->
  top7_make 
```

---

`r chunk_reveal("make_share2", break_type = "auto", widths = c(90, 10), float = "top", title = "## EV in WA state: Shares of makes across counties")`


```{r "make_share2", include=FALSE}
# Shares of top-7 makes across county
ev_data %>%
  group_by(county, make) %>%
  count() %>%
  filter(make %in% top7_make$make) %>%
  group_by(county) %>%
  mutate(county_sum = sum(n), prop = n / county_sum) ->
  make_share_byCounty
```

---

`r chunk_reveal("make_share3", break_type = "auto", widths = c(90, 10), float = "top", title = "## EV in WA state: Shares of makes across counties")`

```{r "make_share3", include=FALSE}
# Coerce `make` into factor and put TESLA as last level
make_share_byCounty %>%
  mutate(make = as.factor(make),
         make = fct_relevel(make, "TESLA", after = Inf)) ->
  make_share_byCounty

# Check levels
head(make_share_byCounty$make)
```

---

`r chunk_reveal("make_share4", break_type = "non_seq", widths = c(50, 50), title = "## EV in WA state: Shares of makes across counties")`

```{r "make_share4", include=FALSE}
ggplot(make_share_byCounty,
       aes(x = prop, 
           y = county, 
           fill = make)) +
  theme_cavis +
  geom_bar(                                 #BREAK2
    position = "stack", stat = "identity",  #BREAK2
    color = "white"                         #BREAK3 
  ) +                                       #BREAK2
  scale_fill_brewer(                        #BREAK4
    type = "qual",                          #BREAK4
    palette = 4 ,                           #BREAK5
    direction = -1                          #BREAK5
  ) +                                       #BREAK4
  scale_x_continuous(                       #BREAK6
    expand = c(0, 0),                       #BREAK6
    labels = label_percent(),               #BREAK7
    sec.axis = dup_axis()                   #BREAK7
  ) +                                       #BREAK6
  theme(aspect.ratio = 1.618) +             #BREAK8
  labs(y = NULL, x = NULL)                  #BREAK9
```

---

`r chunk_reveal("make_share5", break_type = "user", widths = c(50, 50), title = "## EV in WA state: Shares of makes across counties")`

```{r "make_share5", include=FALSE}
ggplot(make_share_byCounty,                 
       aes(x = prop,                        
           y = fct_reorder2(county, make, prop), 
           fill = make)) +                  
  theme_cavis +                             
  geom_bar(                                 
    position = "stack", stat = "identity",  
    color = "white"                         
  ) +                                       
  scale_fill_brewer(                        
    type = "qual",                          
    palette = 4 ,                           
    direction = -1                          
  ) +                                       
  scale_x_continuous(                       
    expand = c(0, 0),                       
    labels = label_percent(),               
    sec.axis = dup_axis()                  
  ) +                                      
  theme(aspect.ratio = 1.618) +         
  labs(y = NULL, x = NULL)        #BREAK

```

---

`r chunk_reveal("ev_income1", break_type = "auto", widths = c(90, 10), float = "top", title = "## EV in WA state: Num. of EVs and median income")`


```{r "ev_income1", include = FALSE}
ev_data %>%
  group_by(county) %>%
  count() %>%
  left_join(county_data, by = "county") %>%
  mutate(n_EV_perhh = n / hh_num * 1000) ->
  ev_by_income 
```

---

`r chunk_reveal("ev_income2", break_type = "auto", widths = c(50, 50), title = "## EV in WA state: Num. of EVs and median income")`

```{r "ev_income2", include = FALSE}
ggplot(ev_by_income,
       aes(x = hh_income, y = n_EV_perhh)) +
  theme_cavis_hgrid +
  geom_point(alpha = 0.75, color = blue) +
  scale_x_continuous(trans = "log10",
                     labels = label_dollar()) +
  scale_y_continuous(trans = "log2") +
  geom_smooth(method = "lm", alpha = 0.2,
              color = blue, fill = blue) +
  labs(y = "N of EVs per 1,000 household",
       x = "Median household income") +
  geom_text_repel(aes(label = county)) 
```

---

`r chunk_reveal("ev_income3", break_type = "auto", widths = c(90, 10), float = "top", title = "## EV in WA state: Num. of EVs and median income")`


```{r "ev_income3", include = FALSE}
# find outliners
lm(n_EV_perhh ~ hh_income, ev_by_income) %>%
  broom::augment(ev_by_income) %>%
  slice_max(order_by = abs(.resid), n = 10) %>%
  pull(county) ->
  outliners
```

---

`r chunk_reveal("ev_income4", break_type = "user", widths = c(50, 50), title = "## EV in WA state: Num. of EVs and median income")`

```{r "ev_income4", include = FALSE}
ggplot(ev_by_income,
       aes(x = hh_income, y = n_EV_perhh)) +
  theme_cavis_hgrid +
  geom_point(alpha = 0.75, color = blue) +
  scale_x_continuous(trans = "log10",
                     labels = label_dollar()) +
  scale_y_continuous(trans = "log2") +
  geom_smooth(method = "lm", alpha = 0.2,
              color = blue, fill = blue) +
  labs(y = "N of EVs per 1,000 household",
       x = "Median household income") + #BREAK
  geom_text_repel(
    data = filter(ev_by_income, 
                  county %in% outliners),
    aes(label = county)
  ) 
```

---

## Some takeaways

- Don't rely on defaults; starts with minimalist aesthetics

--
  
- Understand the consequence of `inherit.aes = TRUE` and how to take advantage/overwrite it 

--
  
- Learn to scale everything (color, fill, x-axis, y-axis, etc)

--
  
- Choose nice colors with `RColorBrewer` 

--
  
- Order things intelligently with `factor`

--
  
- Save graphic output correctly with `ggsave`
